---
title: "ST558 - Project 2"
author: "Sandra Aziz, Simon Weisenhorn"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction




## Loading Libraries
Before we load our data or do any analysis, we will first load the libraries needed throughout the project:
```{r Libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(stringr) # for str_to_title()
library(caret)
library(gbm) #Needed this library to run the training function for boosted tree
```

## Reading in Dataset

Now, we will read in our data set.
```{r Data Import}
# read in the data
data <- read_csv("../data/OnlineNewsPopularity.csv", show_col_types=FALSE)
head(data)
```

```{r Subset Function}
#This function was written to assist in automating data filtering

data_subsetter <- function(channelOfInterest) {
  
  if(!(channelOfInterest %in% c("lifestyle", "entertainment", "bus", "socmed", 
                                "tech", "world"))){
    stop("Error: The channel must be lifestyle, entertainment, bus, socmed, 
         tech, or world")
  }
  #This conditional statement ensures that the user passed one of the possible
  #six data channels, otherwise the function is stopped and an error message is
  #presented
  
  subset <- data %>% filter(eval(parse(text=paste0("data_channel_is_", 
                                                channelOfInterest))) == 1)
  #Subsetting the data to filter the rows based on the given data channel
    
  thinner <- subset[, c(14:19)] #Selecting the 6 data channel columns
  thinner <- thinner[, colSums(thinner != 0) == 0] 
  #Extracting the columns that were not selected by the user input
  drops <- names(thinner) #Selecting the names of irrelevant data channels
  drops <- c(drops, "url") 
  #adding the url column to be removed because it is non-predictive
  data2 <- subset[ , !(names(subset) %in% drops)]
  #removing irrelevant data channels from dataset
  
  channelOfInterest <<- str_to_title(channelOfInterest) 
  #Making global variable out of user supplied channel for later uses
  
  return(data2) 
  #Returning dataset with filtered rows based on the selected channel and 
  #removing the other channels that the user did not supply
}
```

```{r Subset of Data}
channel_data <- data_subsetter("lifestyle") #subsetting for lifestyle channel
head(channel_data) #inspecting subsetted data
```

```{r Train/Test Split, warning=FALSE}
set.seed(558) #setting seed for reproducibility 

trainIndex <- createDataPartition(channel_data$shares, p = 0.7, 
                                  list = FALSE)
#Creating indexes for training data 

channelTrain <- channel_data[trainIndex, ] #Creating training dataset
channelTest <- channel_data[-trainIndex, ] #Creating testing dataset

channelTrain_summary <- channelTrain 
#Creating copy of training data so that summaries can be performed on it below
#This is so that summary variables can be added to the training data without 
#having to add them to the test set

preProcValues <- preProcess(channelTrain, method = c("center", "scale"))
trainTransformed <- predict(preProcValues, channelTrain)
testTransformed <- predict(preProcValues, channelTest)
```



## Some Summaries

```{r Sandra's Summaries}
# summary stats for shares
summary_stats <- summary(channelTrain_summary$shares)
summary_stats

# contingency table for number of videos per url
videos_table <- table(channelTrain_summary$num_videos)
videos_table

# ggplot of number of words in the content vs shares
ggplot(channelTrain_summary, aes(x = n_tokens_content, y = shares)) +
geom_point() +
labs(x = "Number of words in the content", y = "Number of Shares") +
ggtitle("Shares vs. Number of words in the content")

# ggplot of positive word rate vs shares
ggplot(channelTrain_summary, aes(x = rate_positive_words, y = shares)) +
geom_point() +
labs(x = "Positive Word Rate", y = "Number of Shares") +
ggtitle("Shares vs. Positive Word Rate")

# ggplot of negative word rate vs shares
ggplot(channelTrain_summary, aes(x = rate_negative_words, y = shares)) +
geom_point() +
labs(x = "Negative Word Rate", y = "Number of Shares") +
ggtitle("Shares vs. Negative Word Rate")
```


# Simon Weisenhorn's Summaries

```{r Numerical Summaries by Day}
sunday_data <- channelTrain_summary %>% filter(weekday_is_sunday == 1) %>% 
  transmute(day="Sunday", shares=shares)
#Creating new dataset of shares for when the day is Sunday

monday_data <- channelTrain_summary %>% filter(weekday_is_monday == 1) %>% 
  transmute(day="Monday", shares=shares)
#Creating new dataset of shares for when the day is Monday

tuesday_data <- channelTrain_summary %>% filter(weekday_is_tuesday == 1) %>% 
  transmute(day="Tuesday", shares=shares)
#Creating new dataset of shares for when the day is Tuesday

wednesday_data <- channelTrain_summary %>% filter(weekday_is_wednesday == 1) %>% 
  transmute(day="Wednesday", shares=shares)
#Creating new dataset of shares for when the day is Wednesday

thursday_data <- channelTrain_summary %>% filter(weekday_is_thursday == 1) %>% 
  transmute(day="Thursday", shares=shares)
#Creating new dataset of shares for when the day is Thursday

friday_data <- channelTrain_summary %>% filter(weekday_is_friday == 1) %>% 
  transmute(day="Friday", shares=shares)
#Creating new dataset of shares for when the day is Friday

saturday_data <- channelTrain_summary %>% filter(weekday_is_saturday == 1) %>% 
  transmute(day="Saturday", shares=shares)
#Creating new dataset of shares for when the day is Saturday

weekly_share_data <- rbind(sunday_data, monday_data, tuesday_data, 
                           wednesday_data, thursday_data, friday_data, 
                           saturday_data)
#Stacking the rows of the seven previous datasets together with rbind

weekly_share_data %>% group_by(day) %>% summarise(Min = min(shares),
                                              firstQuartile = quantile(shares, 
                                                                       0.25),
                                              Avg=mean(shares),
                                              Med=median(shares),
                                              thirdQuartile = quantile(shares, 
                                                                       0.75),
                                              max = max(shares),
                                              stdDev = sd(shares)) %>%
  arrange(match(day, c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", 
                       "Friday", "Saturday")))
#Creating numerical summary of the amount of shares by day of the week
```

We can inspect the numerical summaries of shares split by the day the article 
was posted. These summaries provide an indication of how the amount of shares
shift as the day of the week changes.

```{r Bar Chart by Day}
weekly_share_summaries <- weekly_share_data %>% group_by(day) %>% 
  summarise(Average = mean(shares), Median = median(shares))
#Creating new variables for the average and median amount of shares by day

weekly_share_summaries$day <- factor(weekly_share_summaries$day, 
                                     levels=c("Sunday", "Monday", "Tuesday", 
                                              "Wednesday", "Thursday", "Friday", 
                                              "Saturday"))
#Reordering the days of the week so that they appear in chronological order in 
#the subsequent graph

weekly_share_summaries <- weekly_share_summaries %>% 
  pivot_longer(cols = c(Average, Median),names_to="statistic", 
               values_to = "value")
#Converting data from wide to long format so that the fill type can be used 
#in the graph below

plot4 <- ggplot(weekly_share_summaries, aes(x=day, y=value, fill=statistic)) 
#Creating the base of the plot

plot4 + geom_bar(stat = "identity", position = "dodge") + 
  #utilizing geom_bar to create a side by side bar graph
  labs(x = "Day of Week", y = "Amount of Shares",
       title = paste0("Bar Chart for Center of Shares by Publish Day for 
    the ",channelOfInterest, " Channel")) +
  #Adding descriptive labels
  theme(plot.title = element_text(hjust = 0.5)) +
  #Centering the title
  scale_fill_discrete(name = "Measure of Center")
  #Renaming the legend for clarity
```

We can inspect the measures of the center for the amount of shares by each day 
an article was posted. If the mean (in red) is taller than the median (in blue)
then the amount of shares is skewed right for that particular day. On the other
hand, if the median (in blue) is taller than the mean (in red) then the amount 
of shares is skewed left for that particular day. Finally, if the mean and 
median are similar heights then the amount of shares are symmetrically 
distributed.

```{r Histogram of Shares by Part of Week, warning=FALSE}
weekend_data <- channelTrain_summary %>% filter(is_weekend == 1) %>% 
  transmute(day="Weekend", shares=shares)
#Creating new dataset of shares for when the article was posted on a weekend

weekday_data <- channelTrain_summary %>% filter(is_weekend == 0) %>% 
  transmute(day="Weekday", shares=shares)
#Creating new dataset of shares for when the article was posted on a weekday

new_weekly_share_data <- rbind(weekend_data, weekday_data)
#Stacking the rows of the seven previous datasets together with rbind

maxxlim <- (quantile(new_weekly_share_data$shares, 0.75) - 
              quantile(new_weekly_share_data$shares, 0.25)) * 1.5
#creating the maximum x value by removing any share values that may be
#considered an outlier so that the histogram is easier to view

plot5 <- ggplot(new_weekly_share_data, aes(x=shares, fill=day)) 
#Creating the base of the plot

plot5 + geom_histogram(data=subset(new_weekly_share_data, day == 'Weekday'), 
                       bins=20, alpha = 0.75) +
  #Adding the first histogram layer for the amount of shares if the article 
  #was posted on a weekday
  geom_histogram(data=subset(new_weekly_share_data, day == 'Weekend'), bins=20, 
                 alpha = 0.75) +
  #Adding the second histogram layer for the amount of shares if the article 
  #was posted on a weekend
  labs(x = "Amount of Shares", y = "Frequency", 
       title = paste0("Histogram of Amount of Shares for the ",channelOfInterest,
                      " Channel")) +
  #Adding descriptive labels
  xlim(0,maxxlim) +
  #limiting the x axis to avoid outliers so that the histogram is viewable
  theme(plot.title = element_text(hjust = 0.5)) +
  #Centering the title
  scale_fill_discrete(name = "Part of Week")
  #Renaming the legend for clarity
```

We can inspect the distributions of the amount of shares based on whether the 
article was posted on a weekday or a weekend. If the distributions are shaped
similarly, then we would have no indication that the distribution of the amount
of shares changes for whether the article was posted on a weekday or a weekend.
However, if one of the distributions are shaped differently than the other, 
such as skewed a certain direction or multi-modal, then we would have reason to
believe that the distribution of the amount of shares changes for whether the 
article was posted on a weekday or a weekend.

```{r Contingency Table of Subjectivity}
med_shares <- median(channelTrain_summary$shares)
#Storing the median amount of shares for a cutoff value

channelTrain_summary <- channelTrain_summary %>% 
  mutate(share_indicator = case_when((shares >= med_shares) ~ "High Shares",
                           (shares < med_shares) ~ "Low Shares"),
#Creating a new variable for indicating when shares are higher than 
#the median amount of all shares for this channel
         subjective_indicator = case_when(global_subjectivity >= 0.5 ~ 
                                            "More Subjective",
                           (global_subjectivity < 0.5) ~ "More Objective"))
#Creating a new variable for indicating when articles are more than half
#subjective or less than half subjective

table(channelTrain_summary$share_indicator, 
      channelTrain_summary$subjective_indicator)
#Creating contingency table for sharing indicator with subjective indicator
```

We can inspect the distributions of the sharing indicator with the subjective
indicator. An article with a high sharing rate is one that has the same amount 
of shares or more than the median amount of shares for all articles in the 
dataset, whereas an article with a low sharing rate is one that has less than 
the median amount of shares for all articles in the dataset. Additionally,
an article is considered more objective if the text subjectivity is less than 
.5, whereas an article is considered more subjective if the text subjectivity
is greater than or equal to .5. Thus, these groups provide insight into 
how sharing is affected when the text subjectivity is changed on a binary scale.

```{r Contingency Table of Images}
channelTrain_summary <- channelTrain_summary %>% 
  mutate(image_indicator = case_when(num_imgs == 0 ~ "No Pictures",
                           num_imgs == 1 ~ "One Picture",
                           num_imgs > 1 ~ "Two or More Pictures",))
#Creating a new variable for indicating when the number of images in the article
#is zero, 1, or more than 1.

table(channelTrain_summary$share_indicator, 
      channelTrain_summary$image_indicator)
#Creating contingency table for sharing indicator with image indicator
```

We can inspect the distributions of the sharing indicator with the image
indicator. An article with a high sharing rate is one that has the same amount 
of shares or more than the median amount of shares for all articles in the 
dataset, whereas an article with a low sharing rate is one that has less than 
the median amount of shares for all articles in the dataset. Additionally,
if the article has no pictures then it is labeled no pictures, or if the article
has a singular picture then it is labeled one picture, or if the article has more
than one picture then it is labeled two or more pictures Thus, these groups 
provide insight into how sharing is affected when the amount of pictures are 
changed on a categorical scale.

```{r Contingency Table of Videos}
channelTrain_summary <- channelTrain_summary %>% 
  mutate(video_indicator = case_when(num_videos == 0 ~ "No Videos",
                           num_videos == 1 ~ "One Video",
                           num_videos > 1 ~ "Two or More Videos",))
#Creating a new variable for indicating when the number of videos in the article
#is zero, 1, or more than 1.

table(channelTrain_summary$share_indicator, 
      channelTrain_summary$video_indicator)
#Creating contingency table for sharing indicator with video indicator
```

We can inspect the distributions of the sharing indicator with the video
indicator. An article with a high sharing rate is one that has the same amount 
of shares or more than the median amount of shares for all articles in the 
dataset, whereas an article with a low sharing rate is one that has less than 
the median amount of shares for all articles in the dataset. Additionally,
if the article has no videos then it is labeled no videos, or if the article
has a singular video then it is labeled one video, or if the article has more
than one video then it is labeled two or more videos Thus, these groups 
provide insight into how sharing is affected when the amount of videos are 
changed on a categorical scale.

```{r Scatterplot for Links}
plot6 <- ggplot(channelTrain_summary, aes(x=num_hrefs, y=shares, 
                                          color=num_self_hrefs)) 
#Creating the base of the plot

plot6 + geom_point() + 
  scale_color_gradient(low="blue", high="red", 
                       name="Number of Links 
Published by 
Mashable") +
  #Here I am adding to the plot to create a scatterplot where the points 
  #are colored by amount of links to other articles published by Mashable
  labs(x = "Number of Links in Article", y = "Amount of Shares",
       title = paste0("Amount of Shares vs Number of Links in an Article for the "
                      ,channelOfInterest," Channel")) +
  #Adding descriptive labels
  theme(plot.title = element_text(hjust = 0.5), title= element_text(size=10), 
        legend.title=element_text(size=10)) 
  #Centering the title and scaling text to fit properly
```

We can inspect the trend of the amount of shares as a function of the number of 
links in an article. If the data points display an upward trend, then we would
find an association between a higher amount of shares given a larger number of 
links in an article. If the data points display a downward trend, then we would 
find an association between a lower amount of shares given a smaller number of 
links in an article. Finally, each datapoint has been colored based on the 
amount of links to other articles published by Mashable. A high amount of 
links to other articles published by Mashable are indicated with red datapoints
and a low amount of links to other articles published by Mashable are indicated
with blue datapoints.

```{r Scatterplot for Keywords}
plot7 <- ggplot(channelTrain_summary, aes(x=num_keywords, y=shares)) 
#Creating the base of the plot

plot7 + geom_point() +
  #Here I am adding to the plot to create a scatterplot 
  labs(x = "Number of Keywords in the Metadata", y="Amount of Shares",
       title = paste0("Amount of Shares vs Number of Keywords in the Metadata of 
  an Article for the ",channelOfInterest," Channel")) +
  #Adding descriptive labels
  theme(plot.title = element_text(hjust = 0.5)) 
  #Centering the title
```

We can inspect the trend of the amount of shares as a function of the number of 
keywords in the metadata. If the data points display an upward trend, then we 
would find an association between a higher amount of shares given a larger 
number of keywords in the metadata for an article. If the data points display a 
downward trend, then we would find an association between a lower amount of 
shares given a smaller number of keywords in the metadata for an article. 

```{r Numerical Summaries by Keywords}
channelTrain_summary %>% group_by(num_keywords) %>% summarise(Min = min(shares),
                                              firstQuartile = quantile(shares, 
                                                                       0.25),
                                              Avg=mean(shares),
                                              Med=median(shares),
                                              thirdQuartile = quantile(shares, 
                                                                       0.75),
                                              max = max(shares),
                                              stdDev = sd(shares)) 
#Creating numerical summary of the amount of shares by number of keywords 
#in the metadata
```

We can inspect the numerical summaries of shares split by the number of keywords 
in the metadata for the article. These summaries provide an indication of how 
the amount of shares shift as the number of keywords in the metadata for the 
article changes.

# Simon's Linear Regression Model

```{r Simon's Linear Fit, warning=FALSE, message=FALSE}
simon_share_lmfit <- train(shares ~ .^2, data = trainTransformed,
method = "lm",
trControl = trainControl(method = "cv", number = 5))
#Fitting linear model with all interactions

summary(simon_share_lmfit)
```

```{r}
data.frame(t(simon_share_lmfit$results))
```

```{r}
data.frame(data = testTransformed$shares, 
                reference = predict(simon_share_lmfit, newdata = testTransformed))
```


# Simon's Boosted Tree Model

```{r message=FALSE, warning=FALSE}

allTuningParameters <- expand.grid(n.trees=c(25,50,100,150,200), 
                                   interaction.depth=c(1,2,3,4), 
                                   shrinkage=0.1, n.minobsinnode = 10)
#Creating dataframe for the tuneGrid arguement using expand.grid

boostTreeFit <- train(shares ~ ., data = trainTransformed,
                      method = "gbm",
                      trControl = trainControl(method = "repeatedcv", 
                                               number = 5, repeats=3),
                      tuneGrid=allTuningParameters,
                      verbose=FALSE)
#Fitting the boosted tree model

boostTreeFit
#viewing the resulting model
```

```{r}
data.frame(data = testTransformed$shares, 
                reference = predict(boostTreeFit, newdata = testTransformed))
```
